---
title: "Inside Prostar core"
subtitle: "A developer's guideline"
author: "Samuel Wieczorek"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  BiocStyle::html_document:
    toc: true
    theme: united
    keep_md: true
    papersize: a4
vignette: >
  %\VignetteIndexEntry{Inside Prostar core}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>")
```


```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```


# Introduction


# Introducing Prostar 2.0


## A collection of modules




# Understanding the source code of workflows



## Initialization of a navigation module instance

At start, 



  * 'Reset': Each process module has a 'Reset' button which allows to set all inputs and plots to their default value if the validation button has not been clicked yet. If the process has several steps, the screen returns to the first one.
  * 'Validate': Each process module has a 'Validate' button which allows to record the new item at the end of the dataset and return the new dataset to Prostar core. After clicking, the screen is not updated: all the inputs keep their current value and the ui is kept to the last step.



The code for the UI is composed of a list a several elements used in the navigation module (mod_navigation.R).



```{r ui_reactive_param, eval=FALSE}
r.nav <- reactiveValues(
    name = "test",
    stepsNames = c("Screen 1", "Screen 2","Screen 3"),
    ll.UI = list( screenStep1 = uiOutput("screen1"),
                  screenStep2 = uiOutput("screen2"),
                  screenStep3 = uiOutput("screen3")),
    isDone =  c(FALSE, FALSE, FALSE),
    mandatory =  c(FALSE, TRUE, FALSE),
    reset = FALSE,
    skip = NULL,
    undo = NULL
    
  )
```  
  
  
* name: the name of the process that is to be managed by the timeline logics,
* steps: A vector of names for the different steps of the process,
* ll.UI: a list in which each item is the UI for one of the screens of the process. The length of this list is
equal to the length of the variables 'steps',
* isDone: A vector of boolean to indicate whether each step has been run by the user. SO, the length of this vector is also equal to the length of the variables 'steps'. By default, all the elements are set to FALSE. They are set to TRUE by some a function in the process source code. The communication between the process and the navigation module is possible because of reactivity. So, if at any moment, an element is set to TRUE by the process, then the corresponding tag is colored in green in the timeline; otherwise, the default color is red if the step is mandatory and orange otherwise,
* isMandatory:
* actions: 




UI
Each process UI is composed of one or more screens which correspond to the different steps of the process. The navigation between each screen is realized by means of buttons placed on the right and the left of the navigation UI (the 'timeline').

The timeline may take different aspects w.r.t the style applied. The principle is to have

Several buttons are available:
* the buttons 'Prev' (on the left) and 'Next' 'on the right' are always part of the timeline. AT the initialization of the process, only the 'Next' button is visible and disabled 

## Logics for processes and datasets

Here are the different rules that exists in Prostar core to deal with the different transactions between data processing modules and the dataset.
The UI of Prostar allows the user to navigate between those two lists so as to rerun a process, fix a mistake in its parameters. But, to guarantee that the workflow stay consistent with the objective/philosophy of Prostar, some rules are implemented.

Each data processing modules and dataset can be seen as a list of respectively P and N items. Let i, p be the indices of respectively the current item in the dataset list and the current item in the process list of the workflow.

The different combination of values for i and p (their relative position) determines the behaviour of datasets and processes regarding the general philosophy of Prostar.

Without any action of the user on the current item nor process, Prostar follows the standard workflow:

* At the beginning of Prostar, i = 1 = N and p = 1
* after the run of any process p, i = N = p -1 (the process has produced a new item in the dataset)

The user may change either the current item in the dataset or the current process to be run. This leads to several situations. 

### Default (standard) behaviour


* The default behaviour one have to see at the end of the execution of the workflow is that it seems to has gone straightforwards in the process list. It means that each process p runs on the item i=N and generates a dataset of N+1 items. If next process to be run is p', then p' > p.


* By default, the behaviour of a module process is to work on the last item of the dataset; the resulting new item is appended to the list of items. Thus, by default, the current indice i is set to the last item of the dataset. For example, if the user starts the workflow with i = 1 (+ offset), p = 1, then at the end of the workflow, one must have i = N, p = P with N = P + 1 (+ offset)


* A process cannot be run more than one time on the same item. To avoid that a process be run on the previous result of itself.

But, the user may navigate differently in the process and dataset lists. Thus, different situations may occur.
Suppose we are after a standard run, then p = P, i = N, i = p + 1


### Reprocess a previous item of the dataset

Suppose i = N and p = P (from the standard workflow). The user can change the current item of the dataset (WHAT'S ITS GOAL ?????)

* i < N (the current item is not the last one of the dataset): Two cases has to be distinguished:
  * p = P ((the current process is the last process of the workflow): xxx
  * p < P (the current process is not the last one of the workflow): xxx

In both cases, one delete all the items of the dataset from i+1 to N then the new item (produced by the process) is appended to the dataset list. 

At the end of this sequence, the current indices must return to their normal values (i = N, p = P). That means that the real workflow has been modified by the user but the final workflow is like the standard one

### 

* a l'affichage de l'UI d'un processus déjà exécuté auparavant:
  * si l'indice courant i est positionné sur le dernier élément du dataset, alors affichage du début du processus,
  * si i est positionné sur un dataset précédent (i < N) (le dataset déjà traité par ce processus), alors on n'affiche que la dernière page du processus


## Nomenclature 


All the files containing code are stored in the directory R.
The modules are prefixed with 'mod_'



## Test modules

Each module in the directory 'R' has a test file, located in 'dev/test_dev'.


# Description of the package Prostar

Prostar is built with the [golem] package.


# Step 1. A simple Shiny App

Prostar is buil with the shiny framework and use the 'navbar page menu' layout. Previous versions of Prostar used other layouts (such as xxx) but the free space on the screen was too small to put all the outputs.

First, let write a simple A


## Workflow files

In each directory of a workflow, there are several type of files (module source code, watch modules code, miscellaneous code).
As those files are dynamically loaded in Prostar (and not at the same time), it is necessary to identify in a unique
manner each file with its function. Two ways are possible to do that:
* build a complex hierarchy of directories in which directory correspond to a specific function,
* name each source code file with a complex name

Actually, Prostar's files structure uses the second point. In the directory of a workflow, the module's files are composed of four strings separated by a '_':
* 'mod' which is the prefix used by the package [golem] to identify the modules,
* the name of the workflow. It seems obvious because we are in the directory of the wf but it is more for the eyes of the developer and for make easiest the reading of the name of the files,
* the name of the process itself.

For the files containing the code to launch the server part of each module, we have prefixed the names by 'watch_'



# UI side

The code for the ui side of Prostar is in the file 'R/app_ui.R'. It consists in two main parts:
* a loading page (div id = 'loading_page'),
* the ui of Prostar ('main_content')

It is based on the navbarPage layout (see xxx) in the shiny package. There are several static menus (and submenus) that are present even if no dataset is loaded:

* **Prostar**:
  * Home: 
  * Global settings:
  * Release notes:
  * Check for updates
* **Data manager**:
  * Open MSnset: ui to open a dataset previously created by the user,
  * Convert: through several steps, converts Excel or CSV files containing quantitative data to an object of class QFeatures,
  which is the format admitted in Prostar,
  * Demo data
  * Reload Prostar
* **Data processing**:
  * Descriptive statistics:
* **Help**:
  * Links:
  * FAQ:
  * Bug report

All of these submenus are implemented as modules : there ui parts are called her and their server part are called from the server side of Prostar (R/app_server.R).





# Server side

The code for the server side of Prostar is in the file 'R/app_server.R'.

## Declaration of global reactive variables


## Call to server modules

The server-side of all the modules which ui are declared in the app_ui.R file are called in this file.


# Load a dataset and corresponding modules

Once a dataset is loaded in prostar (the variable xxx is instanciated), Prostar launches the different modules that are part of the pipeline previously choosen by the user.
The list of modules in that pipeline are stored in the file 'config.R'

## Config.R

This file contains the definition of pipelines, ie the list of processing modules included in the pipeline. For example, here is the definition for the protein pipeline.

```{r ui, eval=FALSE}
pipeline.defs <- list(
  protein = c('Filtering',
              'Normalization',
              'Imputation')
)
}
```

If a developer wants to implement a new pipeline, he has to complete the definition while respecting the nomenclature of the items of the list. This is very important since Prostar maps theses names with source code files.


